---
title: "El Cerrito Police Incident Analysis, 2019--2025"
subtitle: "What the Data Shows About Police Activity, Response, and Geography"
author: "Ira Sharenow"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: false
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 7.5, fig.height = 5, fig.align = "center",
                       fig.crop = FALSE)  # MiKTeX's pdfcrop.exe commonly fails on
                                          # Windows without a working Perl install;
                                          # disabling auto-crop avoids the dependency
                                          # entirely (figures just keep slightly more
                                          # whitespace margin, nothing else changes)

library(tidyverse)
library(readxl)
library(readr)
library(stringr)
library(lubridate)
library(geosphere)
library(ggplot2)
library(ggspatial)
library(sf)
library(ggrepel)
library(knitr)
library(kableExtra)

source_dir <- "D:/Documents/Employment/2026 job search/GitHub/portfolio/Police Department 2026 Git/Police Data/Year Files"
output_dir <- "D:/Documents/Employment/2026 job search/GitHub/portfolio/Police Department 2026 Git/Police Data"
```

```{r load-data}
police <- read_excel(file.path(output_dir, "EC_police_incidents.xlsx")) %>%
  rename_with(~ str_trim(.) %>% str_replace_all(" ", "_") %>% tolower())

police_clean <- police %>%
  filter(!is.na(event_number)) %>%
  distinct() %>%
  mutate(
    address = str_squish(str_replace_all(address, "[\r\n]+", " ")),
    call_for_service = str_squish(call_for_service),
    dt = parse_date_time(received_date, orders = "mdy HM")
  )

incident_types_path <- "D:/Documents/Employment/2025 job search/Project 2025/el-cerrito-police-report/data/combined/types of incidents filtered.csv"

incident_types <- read_csv(incident_types_path, show_col_types = FALSE)[[1]] %>%
  str_trim() %>%
  str_to_lower()

incident_types <- c(incident_types,
  "451 - arson", "20001 - hit and run with injury",
  "20002 - hit and run no injury", "23152 - drunk driving", "602l - trespassing"
)

df_filtered <- police_clean %>%
  filter(str_to_lower(call_for_service) %in% incident_types)

data_start <- min(police_clean$dt, na.rm = TRUE)
data_end <- max(police_clean$dt, na.rm = TRUE)
```

```{r load-geocoding}
BAY_AREA_LAT_MIN <- 37.0
BAY_AREA_LAT_MAX <- 38.5
BAY_AREA_LONG_MIN <- -123.0
BAY_AREA_LONG_MAX <- -121.5
EC_CENTER_LAT <- 37.9161
EC_CENTER_LONG <- -122.3108

original_geocodes <- read_csv(file.path(output_dir, "geocode_checkpoint.csv"), show_col_types = FALSE)
fixed_geocodes <- read_csv(file.path(output_dir, "geocode_checkpoint_v2.csv"), show_col_types = FALSE)

final_geocodes <- original_geocodes %>%
  anti_join(fixed_geocodes, by = "address") %>%
  bind_rows(fixed_geocodes) %>%
  mutate(
    BAD_LATLON_FLAG = is.na(lat) |
      lat < BAY_AREA_LAT_MIN | lat > BAY_AREA_LAT_MAX |
      long < BAY_AREA_LONG_MIN | long > BAY_AREA_LONG_MAX
  )

police_geocoded <- police_clean %>%
  left_join(final_geocodes %>% select(address, lat, long, BAD_LATLON_FLAG), by = "address") %>%
  mutate(BAD_LATLON_FLAG = if_else(is.na(address), TRUE, BAD_LATLON_FLAG))

incidents_for_proximity <- police_geocoded %>% filter(!BAD_LATLON_FLAG)

police_total <- incidents_for_proximity %>%
  filter(str_to_lower(call_for_service) %in% incident_types)

# Second-round outlier correction (addresses that landed in the right
# general region but far from El Cerrito itself)
outliers_result <- read_csv(file.path(output_dir, "geocode_outliers_v3.csv"), show_col_types = FALSE)

KNOWN_DISTANT_CITIES <- "SAN FRANCISCO|PIEDMONT|OAKLAND|PINOLE|STOCKTON|PETALUMA|SAN LEANDRO|CONCORD|ANTIOCH"

outliers_final <- outliers_result %>%
  mutate(
    dist_from_ec_mi = distHaversine(cbind(long, lat), c(EC_CENTER_LONG, EC_CENTER_LAT)) / 1609.34,
    exclusion_reason = case_when(
      is.na(lat) ~ "unresolved - no match",
      dist_from_ec_mi <= 5 ~ "usable",
      str_detect(address, KNOWN_DISTANT_CITIES) ~ "confirmed different city",
      TRUE ~ "unresolved - shorthand matched wrong location"
    )
  )

usable_outliers <- outliers_final %>%
  filter(exclusion_reason == "usable") %>%
  select(address, lat, long)

police_total_corrected <- police_total %>%
  rows_update(usable_outliers, by = "address") %>%
  mutate(dist_from_ec_mi = distHaversine(cbind(long, lat), c(EC_CENTER_LONG, EC_CENTER_LAT)) / 1609.34) %>%
  filter(dist_from_ec_mi <= 5)
```

```{r landmarks}
bay_area_extent <- paste(BAY_AREA_LONG_MIN, BAY_AREA_LAT_MIN, BAY_AREA_LONG_MAX, BAY_AREA_LAT_MAX, sep = ",")

landmarks <- tribble(
  ~landmark, ~address,
  "El Cerrito Plaza BART", "6699 Fairmount Avenue, El Cerrito, CA 94530",
  "El Cerrito Del Norte BART", "6400 Cutting Blvd, El Cerrito, CA 94530",
  "Castro Park Pickleball", "1420 Norvell St, El Cerrito, CA 94530",
  "El Cerrito Community Center", "7007 Moeser Ln, El Cerrito, CA 94530",
  "El Cerrito High School", "540 Ashbury Ave, El Cerrito, CA 94530",
  "Korematsu Middle School", "7125 Donal Ave, El Cerrito, CA 94530",
  "Harding Elementary School", "7230 Fairmount Ave, El Cerrito, CA 94530",
  "Madera Elementary School", "8500 Madera Dr, El Cerrito, CA 94530",
  "Current Library", "6510 Stockton Ave, El Cerrito, CA 94530"
)

landmarks_geocoded <- landmarks %>%
  tidygeocoder::geocode(address, method = "arcgis", lat = lat, long = long,
                        custom_query = list(searchExtent = bay_area_extent))

# Korematsu: ArcGIS matches the correct address text with full confidence,
# but its underlying reference data still points at the school's pre-2015
# location. Hand-corrected against a coordinate read directly off a current
# map. See the Technical Appendix for how this was caught.
landmarks_geocoded <- landmarks_geocoded %>%
  mutate(
    lat = if_else(landmark == "Korematsu Middle School", 37.9208100248211, lat),
    long = if_else(landmark == "Korematsu Middle School", -122.30584773399178, long)
  ) %>%
  mutate(landmark = case_when(
    landmark == "El Cerrito Del Norte BART"   ~ "Del Norte BART",
    landmark == "Madera Elementary School"    ~ "Madera",
    landmark == "Korematsu Middle School"     ~ "Korematsu",
    landmark == "El Cerrito Community Center" ~ "EC Community Center",
    landmark == "El Cerrito High School"      ~ "ECHS",
    landmark == "Harding Elementary School"   ~ "Harding",
    landmark == "El Cerrito Plaza BART"       ~ "EC Plaza BART",
    landmark == "Castro Park Pickleball"      ~ "Castro Park",
    landmark == "Current Library"             ~ "Library",
    TRUE ~ landmark
  ))

RADIUS_FT <- 500
RADIUS_M <- RADIUS_FT * 0.3048
```

\newpage

# Introduction

This report analyzes every police incident logged by the El Cerrito Police
Department from **`r format(data_start, "%B %d, %Y")` through `r format(data_end, "%B %d, %Y")`** --
`r format(nrow(police_clean), big.mark = ",")` records in total. It looks at
four questions residents might reasonably ask about their local police
department: what officers actually spend their time on, where activity is
concentrated, how the pattern has changed over time, and how quickly the
department responds when something happens.

This is not a crime-statistics report in the traditional sense. It treats
the department's own incident log as a dataset worth understanding on its
own terms -- most of what gets logged isn't a crime at all, and that turns
out to be one of the most interesting findings in the whole analysis.

A detailed technical appendix at the end of this report documents exactly
how the underlying data was assembled, cleaned, and verified, including
several real errors that were found and fixed along the way. That level of
detail is there for anyone who wants to check the work; the sections below
focus on what the data actually shows.

\newpage

# What Do El Cerrito Police Actually Do?

```{r workload}
workload_summary <- police_clean %>%
  count(call_for_service, sort = TRUE) %>%
  mutate(
    in_crime_scope = str_to_lower(call_for_service) %in% incident_types,
    pct_of_total = round(100 * n / sum(n), 1)
  )

crime_scope_total <- workload_summary %>%
  summarize(total = sum(n), crime_scope = sum(n[in_crime_scope])) %>%
  mutate(pct = round(100 * crime_scope / total, 1))
```

Before looking at where or when incidents happen, it's worth asking a more
basic question: what does the department actually spend its time on? The
answer is not what most people would guess.

Of the `r format(crime_scope_total$total, big.mark = ",")` incidents in this
dataset, only **`r crime_scope_total$pct`%** fall into a category most
people would call a crime -- theft, burglary, assault, vandalism, and
similar. The remaining `r 100 - crime_scope_total$pct`% is patrol and
administrative work: security checks, traffic stops, alarm responses, 911
disconnects, welfare checks, and calls that resolve before an officer
arrives at all.

```{r workload-table}
workload_summary %>%
  slice_head(n = 10) %>%
  mutate(call_for_service = str_to_title(call_for_service)) %>%
  select(`Call Type` = call_for_service, Count = n, `% of Total` = pct_of_total) %>%
  kable(caption = "Ten most common call types, all years combined") %>%
  kable_styling(latex_options = "hold_position")
```

The single most common thing El Cerrito police respond to is a security
check -- an officer proactively checking that a building or area is secure,
not responding to a report of anything wrong. Traffic stops are second.
Together with alarm calls, 911 disconnects, and follow-up visits, these
few categories alone account for close to half of everything the
department logs. This is worth keeping in mind throughout the rest of this
report: a "hot" location on the maps that follow reflects concentrated
*crime*, specifically, not general police presence.

\newpage

# Where Does Activity Happen?

```{r general-heatmap, fig.cap="Density of crime-scope incidents, all categories, relative to nine community landmarks"}
EC_MAP_LAT_MIN <- 37.885
EC_MAP_LAT_MAX <- 37.94
EC_MAP_LONG_MIN <- -122.325
EC_MAP_LONG_MAX <- -122.275

ggplot() +
  annotation_map_tile(type = "osm", zoomin = 0) +
  stat_density_2d(
    data = police_total_corrected,
    aes(x = long, y = lat, fill = after_stat(level)),
    geom = "polygon", contour = TRUE, alpha = 0.5, bins = 25
  ) +
  scale_fill_gradient(low = "yellow", high = "red", breaks = range,
                       labels = c("Fewer incidents", "More incidents")) +
  geom_point(data = landmarks_geocoded, aes(x = long, y = lat),
             color = "black", size = 2.5, shape = 17) +
  geom_label_repel(data = landmarks_geocoded, aes(x = long, y = lat, label = landmark),
                    size = 2.8, fontface = "bold", label.padding = 0.15,
                    label.size = 0, fill = alpha("white", 0.75),
                    min.segment.length = 0, max.overlaps = Inf) +
  coord_sf(xlim = c(EC_MAP_LONG_MIN, EC_MAP_LONG_MAX),
           ylim = c(EC_MAP_LAT_MIN, EC_MAP_LAT_MAX), crs = 4326) +
  labs(title = "El Cerrito Crime-Scope Incident Density", fill = NULL) +
  theme_minimal() +
  theme(axis.title = element_blank())
```

Crime-scope incidents concentrate heavily along the San Pablo Avenue
corridor, El Cerrito's main commercial spine, with the two clearest
hotspots near El Cerrito Plaza BART and the cluster around the Library,
El Cerrito High School, and Harding Elementary. Activity drops off sharply
moving east into the hillside residential neighborhoods, where Madera and
Korematsu sit.

```{r landmark-table}
landmark_counts <- landmarks_geocoded %>%
  select(landmark, land_lat = lat, land_long = long) %>%
  rowwise() %>%
  mutate(n_within_500ft = sum(
    distHaversine(cbind(incidents_for_proximity$long, incidents_for_proximity$lat),
                  c(land_long, land_lat)) <= RADIUS_M
  )) %>%
  ungroup() %>%
  arrange(desc(n_within_500ft))

landmark_counts %>%
  select(Landmark = landmark, `Incidents within 500 ft` = n_within_500ft) %>%
  kable(caption = "All police activity within 500 feet of each landmark (all call types, not just crime-scope)") %>%
  kable_styling(latex_options = "hold_position")
```

Table 2 counts *all* police activity, not just crime, within 500 feet of
nine community landmarks. One entry deserves a specific caveat: El Cerrito
Del Norte BART sits directly on the city's border with Richmond and
includes its own dedicated BART Police substation. Incidents on BART
property or across the Richmond city line are handled by a different
department entirely and are not part of this dataset -- so this number
reflects only nearby off-station activity that happened to fall on the El
Cerrito side, not the full picture of what happens at that station.

## Vehicle Crime and Dangerous Driving

```{r vehicle-subsets}
vehicle_crime_types <- c("459a - auto burglary", "10851 - motor vehicle theft",
                          "10851r - recovered stolen vehicle", "215 - car jacking",
                          "23110 - throwing objects at a vehicle", "10852 - vehcile parts theft")
dangerous_driving_types <- c("20001 - hit and run with injury", "20002 - hit and run no injury",
                              "23152 - drunk driving")

police_vehicle_crime <- police_total_corrected %>%
  filter(str_to_lower(call_for_service) %in% vehicle_crime_types)
police_dangerous_driving <- police_total_corrected %>%
  filter(str_to_lower(call_for_service) %in% dangerous_driving_types)
police_other_crime <- police_total_corrected %>%
  filter(!(str_to_lower(call_for_service) %in% c(vehicle_crime_types, dangerous_driving_types)))

compute_landmark_counts <- function(incident_df, label) {
  landmarks_geocoded %>%
    select(landmark, land_lat = lat, land_long = long) %>%
    rowwise() %>%
    mutate(n_within_500ft = sum(
      distHaversine(cbind(incident_df$long, incident_df$lat), c(land_long, land_lat)) <= RADIUS_M
    )) %>%
    ungroup() %>%
    select(landmark, {{label}} := n_within_500ft)
}

landmark_comparison <- compute_landmark_counts(police_vehicle_crime, "vehicle_crime") %>%
  left_join(compute_landmark_counts(police_dangerous_driving, "dangerous_driving"), by = "landmark") %>%
  left_join(compute_landmark_counts(police_other_crime, "other_crime"), by = "landmark") %>%
  mutate(total = vehicle_crime + dangerous_driving + other_crime) %>%
  arrange(desc(total))
```

Vehicle crime (auto burglary, motor vehicle theft, carjacking) and
dangerous driving (hit and run, drunk driving) are broken out separately
below, since they represent genuinely different kinds of risk -- property
crime against a parked car versus danger from a car in motion.

```{r landmark-comparison-table}
landmark_comparison %>%
  select(Landmark = landmark, `Vehicle Crime` = vehicle_crime,
         `Dangerous Driving` = dangerous_driving, `All Other Crime` = other_crime,
         Total = total) %>%
  kable(caption = "Crime-scope incidents within 500 ft, by landmark and category") %>%
  kable_styling(latex_options = "hold_position") %>%
  footnote(general = "\"All Other Crime\" is every crime-scope category not already broken out as Vehicle Crime or Dangerous Driving -- for example assault, robbery, burglary not involving a vehicle, theft, and vandalism.",
           general_title = "Note: ", footnote_as_chunk = TRUE)
```

```{r vehicle-heatmap, fig.cap="Density of vehicle-crime incidents specifically"}
ggplot() +
  annotation_map_tile(type = "osm", zoomin = 0) +
  stat_density_2d(
    data = police_vehicle_crime,
    aes(x = long, y = lat, fill = after_stat(level)),
    geom = "polygon", contour = TRUE, alpha = 0.5, bins = 15
  ) +
  scale_fill_gradient(low = "yellow", high = "red", breaks = range,
                       labels = c("Fewer incidents", "More incidents")) +
  geom_point(data = landmarks_geocoded, aes(x = long, y = lat),
             color = "black", size = 2.5, shape = 17) +
  geom_label_repel(data = landmarks_geocoded, aes(x = long, y = lat, label = landmark),
                    size = 2.8, fontface = "bold", label.padding = 0.15,
                    label.size = 0, fill = alpha("white", 0.75),
                    min.segment.length = 0, max.overlaps = Inf) +
  coord_sf(xlim = c(EC_MAP_LONG_MIN, EC_MAP_LONG_MAX),
           ylim = c(EC_MAP_LAT_MIN, EC_MAP_LAT_MAX), crs = 4326) +
  labs(title = "El Cerrito Vehicle Crime Density", fill = NULL) +
  theme_minimal() +
  theme(axis.title = element_blank())
```

EC Plaza BART and the Community Center lead in both vehicle crime and
dangerous driving specifically, not just overall activity -- a consistent
pattern, not an artifact of one dominant category.

\newpage

# When Does It Happen?

```{r yearly-trend, fig.cap="Crime-scope incidents by year"}
yearly_totals <- police_clean %>%
  filter(str_to_lower(call_for_service) %in% incident_types) %>%
  mutate(year = year(dt)) %>%
  count(year, name = "incidents") %>%
  mutate(
    year_label = if_else(year == 2025, "2025\n(partial year)", as.character(year)),
    bar_fill = if_else(year == 2025, "2025", "full year")
  ) %>%
  mutate(year_label = fct_reorder(year_label, year))

ggplot(yearly_totals, aes(x = year_label, y = incidents, fill = bar_fill)) +
  geom_col() +
  geom_text(aes(label = incidents), vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("full year" = "steelblue", "2025" = "gray60"), guide = "none") +
  labs(x = NULL, y = "Incidents", title = "Crime-Scope Incidents by Year") +
  theme_minimal()
```

Incident counts dropped noticeably during 2020--2022, bottoming out in
2022, before recovering to their highest levels of the period in 2023 and
2024. This report does not attempt to explain *why* -- police staffing
levels, changes in what gets reported, and actual changes in activity
could all play a role, and disentangling them is beyond what this dataset
can show on its own.

2025 is shown through June 30 only and should not be compared directly to
the full-year totals above. On a fair like-for-like basis -- January
through June of every year -- 2025 is actually the **busiest first half of
any year in this seven-year period**, continuing the recovery rather than
reversing it.

```{r monthly-trend, fig.cap="Average monthly pattern, 2019-2024 (2025 excluded as a partial year)"}
monthly_totals <- police_clean %>%
  filter(str_to_lower(call_for_service) %in% incident_types, year(dt) %in% 2019:2024) %>%
  mutate(month = month(dt, label = TRUE)) %>%
  count(month, name = "incidents")

avg_month <- mean(monthly_totals$incidents)

ggplot(monthly_totals, aes(x = month, y = incidents)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = avg_month, linetype = "dashed", color = "darkred") +
  labs(x = NULL, y = "Incidents (2019-2024 total)",
       title = "Crime-Scope Incidents by Month") +
  theme_minimal()
```

October stands out from the rest of the year -- but not for the reason it
might first appear. Breaking the October total down by category shows it's
driven almost entirely by parking enforcement, not a broad rise in crime.
Parking violations get their own section below, since they turned out to
be one of the more striking trends in the whole dataset.

\newpage

# Parking Enforcement: A City on the Move

```{r parking-yearly, fig.cap="Parking violations by year"}
parking_yearly <- police_clean %>%
  filter(call_for_service == "PARKER - PARKING VIOLATION") %>%
  mutate(year = year(dt)) %>%
  count(year, name = "parking_violations") %>%
  mutate(
    year_label = if_else(year == 2025, "2025\n(partial year)", as.character(year)),
    bar_fill = if_else(year == 2025, "2025", "full year")
  ) %>%
  mutate(year_label = fct_reorder(year_label, year))

ggplot(parking_yearly, aes(x = year_label, y = parking_violations, fill = bar_fill)) +
  geom_col() +
  geom_text(aes(label = parking_violations), vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("full year" = "darkorange", "2025" = "wheat3"), guide = "none") +
  labs(x = NULL, y = "Parking Violations", title = "Parking Violations by Year") +
  theme_minimal()
```

Parking violations more than tripled over the period covered by this
report, from `r parking_yearly$parking_violations[parking_yearly$year == 2019]`
in 2019 to `r parking_yearly$parking_violations[parking_yearly$year == 2024]`
in 2024. The growth is not gradual -- it happened in two distinct jumps
rather than a steady climb, which points to specific operational changes
at identifiable points in time rather than organic drift. A closer look at
the timing of those changes is a natural next step for a future update to
this report.

```{r parking-monthly, fig.cap="Average monthly parking-violation pattern, 2019-2024"}
parking_monthly <- police_clean %>%
  filter(call_for_service == "PARKER - PARKING VIOLATION", year(dt) %in% 2019:2024) %>%
  mutate(month = month(dt, label = TRUE)) %>%
  count(month, name = "parking_violations")

ggplot(parking_monthly, aes(x = month, y = parking_violations)) +
  geom_col(fill = "darkorange") +
  labs(x = NULL, y = "Parking Violations (2019-2024 total)",
       title = "Parking Violations by Month") +
  theme_minimal()
```

## Is Enforcement Concentrating Near El Cerrito Plaza?

```{r parking-plaza}
ec_plaza_bart <- landmarks_geocoded %>% filter(landmark == "EC Plaza BART")

parking_geocoded <- police_clean %>%
  filter(call_for_service == "PARKER - PARKING VIOLATION") %>%
  left_join(final_geocodes %>% select(address, lat, long, BAD_LATLON_FLAG), by = "address") %>%
  filter(!BAD_LATLON_FLAG, !is.na(lat)) %>%
  mutate(dist_to_plaza_ft = distHaversine(cbind(long, lat), c(ec_plaza_bart$long, ec_plaza_bart$lat)) / 0.3048)

parking_plaza_by_year <- parking_geocoded %>%
  mutate(year = year(dt), within_1000ft = dist_to_plaza_ft <= 1000) %>%
  group_by(year) %>%
  summarize(
    total = n(),
    near_plaza = sum(within_1000ft),
    pct_near_plaza = round(100 * near_plaza / total, 1)
  )
```

With talk locally of a parking district near El Cerrito Plaza, it's worth
asking directly: is a growing share of citywide parking enforcement
actually concentrating there? To check, every geocoded parking violation
was measured against its distance from the Plaza BART station.

```{r parking-plaza-table}
parking_plaza_by_year %>%
  select(Year = year, `Citywide Total` = total,
         `Within 1,000 ft of Plaza` = near_plaza,
         `% of Citywide Total` = pct_near_plaza) %>%
  kable(caption = "Parking violations near El Cerrito Plaza, by year") %>%
  kable_styling(latex_options = "hold_position") %>%
  footnote(general = "2025 reflects only January through June (a partial year). The percentage column is not affected by this, since both the citywide total and the near-Plaza count are equally partial.",
           general_title = "Note: ", footnote_as_chunk = TRUE)
```

The answer is no -- at least not so far. The *share* of citywide parking
tickets written within 1,000 feet of the Plaza has not risen over this
period; if anything, it is somewhat lower in 2025 than it was in 2019.
The raw number of tickets near the Plaza has grown, but that simply
reflects the citywide tripling described above, not extra concentration
in that specific area. Whatever is driving the parking-district
conversation, this dataset does not show enforcement disproportionately
shifting toward the Plaza.

\newpage

# How Fast Do Police Respond?

```{r response-time}
response_data <- police_clean %>%
  filter(str_to_lower(call_for_service) %in% incident_types) %>%
  mutate(
    arrived_dt = parse_date_time(first_unit_arrived_time, orders = "mdy HM"),
    response_min = as.numeric(difftime(arrived_dt, dt, units = "mins"))
  ) %>%
  filter(!is.na(response_min), response_min >= 0, response_min < 1440)

response_summary <- response_data %>%
  summarize(
    median_min = median(response_min),
    p90_min = round(quantile(response_min, 0.9), 1)
  )

urgent_types <- c("245 - assault with a deadly weapon", "211 - robbery",
                   "1071 - shooting (actual victim)", "sfrmc - shots fired",
                   "215 - car jacking")

urgent_summary <- response_data %>%
  filter(str_to_lower(call_for_service) %in% urgent_types) %>%
  summarize(median_min = median(response_min))
```

The department's median response time across all in-scope incidents is
**`r response_summary$median_min` minutes**, and 90% of responses arrive
within `r response_summary$p90_min` minutes. For calls involving active,
in-progress danger specifically -- shots fired, an armed assault, a
robbery, a carjacking, a shooting with an actual victim -- the median
response is **`r urgent_summary$median_min` minutes**, consistently fast
across every one of those categories.

A word of caution for anyone comparing response times across years: a
naive year-by-year average makes it look like response times dramatically
improved, dropping from around 7 minutes in 2019 to 2 minutes in 2025.
That apparent improvement is misleading. It's driven almost entirely by
the huge growth in parking enforcement described above -- writing a
parking ticket involves no dispatch or travel time, so as parking
violations became a larger share of everything logged, the citywide
average mechanically dropped without response to genuine emergencies
actually getting any faster. Looking only at incidents that involve a real
dispatch and response, the department's response time has been
**stable, not dramatically changing, across the entire seven-year
period** -- a consistent, reliable finding rather than a flashy but
misleading one.

\newpage

# Conclusion

Four findings stand out from this analysis. First, most of what El
Cerrito police do isn't crime response at all -- fewer than three calls in
ten fall into a category most residents would call a crime; the rest is
patrol, traffic enforcement, and administrative work. Second, crime-scope
activity concentrates heavily along the San Pablo Avenue corridor, with
the sharpest drop-off in the hillside residential areas -- and when a
landmark's number looks surprisingly low, as with Del Norte BART, it's
often because of a real jurisdictional boundary, not because nothing
happens there. Third, parking enforcement has more than tripled citywide
since 2019, but that growth is broad-based rather than concentrated near
El Cerrito Plaza specifically, at least through the period this report
covers. Fourth, the department's response to genuinely dangerous,
in-progress incidents is fast and consistent, holding steady across the
whole period studied, even as the overall mix of what gets logged has
shifted substantially.

This report will be updated with a second analysis covering July 2025
through June 2026 once that data becomes available.

\newpage

# Technical Appendix

This appendix summarizes how the underlying dataset was built, cleaned,
and verified. It is intentionally more detailed than a typical report
appendix, because several real data-quality problems were found and fixed
during this project, and the goal here is to show that work rather than
hide it.

## Data Source

Source data was 13 half-year PDF reports from the El Cerrito Police
Department, covering `r format(data_start, "%B %d, %Y")` through
`r format(data_end, "%B %d, %Y")`. Two extraction defects were found and
corrected in the initial PDF-to-spreadsheet conversion: a small share of
addresses were truncated when they wrapped across a line in the source
PDF, and the first one or two rows on nearly every PDF page were initially
dropped entirely. Both were caught by independently re-extracting the data
and diffing the two versions against each other and against the raw PDF
text directly, then verified file-by-file across all 13 source documents.

After combining and deduplicating (including 376 rows that were genuine
duplicates spanning a source-file boundary in mid-2020), the working
dataset contains `r format(nrow(police_clean), big.mark = ",")` unique
incidents.

## Incident Categories

This report distinguishes between the full incident log and a
crime-scope subset of `r format(nrow(df_filtered), big.mark = ",")`
incidents, built from a 52-category list of genuine crime types (excluding
administrative categories like security checks and traffic stops, and
also excluding financial crimes such as identity theft, since those are
typically logged at the victim's home address rather than a crime scene
and would distort a geographic analysis).

## Geocoding

Street addresses were converted to map coordinates using the ArcGIS
World Geocoding Service. An initial geocoding pass revealed that,
without a geographic constraint, the service would sometimes match an
ambiguous address (an intersection referenced by a local abbreviation, or
a generic place name) to a same-named location anywhere in the world,
with full confidence -- one address matched the country of Panama, another
matched a park in Australia. Adding an explicit geographic boundary to the
geocoding request and re-processing the affected addresses resolved the
large majority of these cases; a final distance-based check excluded any
remaining addresses that could not be reliably placed. In total,
`r format(nrow(incidents_for_proximity), big.mark = ",")` of
`r format(nrow(police_clean), big.mark = ",")` incidents
(`r round(100 * nrow(incidents_for_proximity) / nrow(police_clean), 1)`%)
were successfully and reliably geocoded.

Community landmarks used in the geographic analysis were verified by hand
against independent sources rather than accepted on the geocoding
service's confidence score alone. This caught one landmark (Korematsu
Middle School) where the service returned a maximum-confidence match on
the correct current address, but an outdated underlying coordinate that
still pointed to the school's location prior to a relocation roughly a
decade ago -- corrected manually after independent verification.

## Limitations

This dataset reflects only incidents logged by the El Cerrito Police
Department within city jurisdiction. Activity occurring on BART property,
across the Richmond city line, or handled by a different agency is not
captured here, which is worth keeping in mind when interpreting any single
landmark's numbers, particularly El Cerrito Del Norte BART. Response-time
figures are based on the subset of incidents with complete timestamp data
(88.3% of crime-scope incidents); a small number of long-delay outliers,
consistent with reports made well after an incident occurred rather than
an active emergency, were identified and are described in supporting
project documentation.
